---
title: "Support Vector Machines"
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
date: "2024-11-24"
---
SVM Analysis of Drug Response Data

---
title: "SVM Analysis - Drug Response Prediction"
output:
  html_document: default
  pdf_document:
    latex_engine: xelatex
date: "2024-11-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(e1071)
library(caret)
library(tidyverse)
library(skimr)
```

### Data Loading and Initial Exploration

```{r}
drug_data <- read.csv('https://raw.githubusercontent.com/Kingtilon1/MachineLearning-BigData/refs/heads/main/DecisionTree/drug200.csv')

head(drug_data)
str(drug_data)
skim(drug_data)
```

### Data Preprocessing

```{r}
drug_data$Sex <- as.factor(drug_data$Sex)
drug_data$BP <- as.factor(drug_data$BP)
drug_data$Cholesterol <- as.factor(drug_data$Cholesterol)
drug_data$Drug <- as.factor(drug_data$Drug)

set.seed(123)
train_index <- createDataPartition(drug_data$Drug, p = 0.8, list = FALSE)
train_data <- drug_data[train_index, ]
test_data <- drug_data[-train_index, ]

preproc <- preProcess(train_data[, c("Age", "Na_to_K")], method = c("center", "scale"))
train_data_scaled <- predict(preproc, train_data)
test_data_scaled <- predict(preproc, test_data)
```

### SVM Model with Linear Kernel

```{r}
svm_linear <- svm(Drug ~ ., data = train_data_scaled, kernel = "linear", cost = 1)

pred_linear <- predict(svm_linear, test_data_scaled)
conf_matrix_linear <- confusionMatrix(pred_linear, test_data_scaled$Drug)
print(conf_matrix_linear)
```

### SVM Model with Radial Kernel

```{r}
svm_radial <- svm(Drug ~ ., data = train_data_scaled, kernel = "radial", cost = 1)

pred_radial <- predict(svm_radial, test_data_scaled)
conf_matrix_radial <- confusionMatrix(pred_radial, test_data_scaled$Drug)
print(conf_matrix_radial)
```

### Tuning SVM Parameters

```{r}
tuning_grid <- expand.grid(
  cost = c(0.1, 1, 10),
  gamma = c(0.1, 1, 10)
)

svm_tune <- tune.svm(
  Drug ~ ., 
  data = train_data_scaled,
  kernel = "radial",
  cost = c(0.1, 1, 10),
  gamma = c(0.1, 1, 10)
)
print(svm_tune$best.parameters)

svm_final <- svm(
  Drug ~ .,
  data = train_data_scaled,
  kernel = "radial",
  cost = svm_tune$best.parameters$cost,
  gamma = svm_tune$best.parameters$gamma
)

pred_final <- predict(svm_final, test_data_scaled)
conf_matrix_final <- confusionMatrix(pred_final, test_data_scaled$Drug)
print(conf_matrix_final)
```

### Variable Importance Analysis

```{r}
importance <- data.frame(
  Feature = names(train_data_scaled)[-which(names(train_data_scaled) == "Drug")],
  Importance = 0
)

for(feature in importance$Feature) {
  test_permuted <- test_data_scaled
  test_permuted[,feature] <- sample(test_permuted[,feature])
  
  pred_permuted <- predict(svm_final, test_permuted)
  
  importance$Importance[importance$Feature == feature] <- 
    mean(pred_final == test_data_scaled$Drug) - mean(pred_permuted == test_data_scaled$Drug)
}

importance <- importance[order(-importance$Importance),]
ggplot(importance, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Feature Importance in SVM Model",
       x = "Features",
       y = "Importance (Accuracy Drop)")
```

### Comparison with Decision Tree Results

```{r}
results_comparison <- data.frame(
  Model = c("Decision Tree", "SVM Linear", "SVM Radial", "SVM Tuned"),
  Accuracy = c(0.97,
               conf_matrix_linear$overall["Accuracy"],
               conf_matrix_radial$overall["Accuracy"],
               conf_matrix_final$overall["Accuracy"]),
  Kappa = c(0.96, 
            conf_matrix_linear$overall["Kappa"],
            conf_matrix_radial$overall["Kappa"],
            conf_matrix_final$overall["Kappa"])
)

print(results_comparison)
```